Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-12-02 paper2_JNLS/paper.tex
NM

\begin{abstract} 
Recent laboratory experiments of \boetal demonstrate that an abrupt change in bottom topography can trigger anomalous statistics in randomized surface waves. Motivated by these observations, \maetal developed a theoretical framework, based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) system, that successfully captures key qualitative features of the experiments, including the robust emergence of anomalous statistics and heightened skewness in the outgoing wavefield. Here, we extend these parallel experimental and modeling efforts with several new findings that have resulted from a synergetic interaction between the two. By precisely relating model parameters to physical ones, we calibrate the model inverse temperature to the specific conditions present in the experiments, thereby permitting a quantitative comparison. We find theoretically predicted distributions of surface displacement to match the experimental measurements with surprising quantitive detail. Prompted by the presence of surface slope in the TKdV Hamiltonian, we present new experimental measurements on surface slope statistics and compare them to model predictions. Analysis of some deterministic trajectories of TKdV elucidates the experimental length and time scales required for the statistical transition to a skewed state. Finally, the theory predicts a peculiar relationship between the outgoing displacement skewness and the change in slope variance, specifically how their ratio depends on the wave amplitude and depth ratio. New experimental measurements confirm this prediction in spectacular fashion.
%New experimental measurements provide convincing evidence for this prediction.
\end{abstract}

@@

Our particular focus is the laboratory experiments of \boetal~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing over the step, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, \maetal \cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda \& Qi (2019) \cite{majdaqi2019} analyzed more severely truncated systems --- as low as two interacting modes --- that exhibit a statistical phase transition to anomalous statistics and the creation of extreme events while enjoying a more tractable structure. More recently, Qi \& Majda (2019) \cite{MachineLearning2019} demonstrated the capability of machine learning strategies to accurately predict these extreme events. That work uses a deep neural network coupled with a judicious choice of the entropy loss function that emphasizes the dominant structures of the turbulent field \cite{MachineLearning2019}. It should be noted that, in these studies, extreme events are represented by strong skewness of the wave-field. They occur intermittently and on a relatively frequent basis, which is in contrast to the simpler situation of isolated, rare events \cite{guth2019machine}.

The purpose of the present manuscript is to provide a more comprehensive treatment of the {\em combined} experimental and theoretical efforts of \boetal~\cite{bolles2019} and \maetal~\cite{majda2019, majdaqi2019}. Special emphasis is placed on the synergy between experiment and theory -- a cooperative strategy with a proven record of success \cite{camassa2012stratified, ristroph2012, ganedi2018equilibrium}. In Section \ref{experiments} we detail the laboratory experiments of \boetal \cite{bolles2019} and summarize previously reported findings on anomalous statistics of the surface displacement. We also present new experimental data on surface {\em slope} statistics --- a line of inquiry motivated by the theoretical advancements of \maetal \cite{majda2019}, specifically the central role played by the slope in the Hamiltonian structure of TKdV.  Section \ref{theory} discusses the TKdV theoretical framework, including both the deterministic and the statistical mechanics perspectives. In this section, we flesh out details of the non-dimensionalization, which, for brevity, were only briefly discussed in previous work. This exercise provides a more thorough link between model and experimental parameters, ultimately facilitating a closer comparison between the two. We then discuss TKdV as a deterministic dynamical system, approachable from the viewpoint of statistical mechanics. The later perspective is based on the Hamiltonian structure and novel ensemble distributions that incorporate canonical and micro-canonical aspects \cite{abramov2003}. Section \ref{results} presents a systematic comparison between experiments and theory, focusing on both surface displacement and surface slope. We find the statistical distributions of these two quantities to agree well across experiments and theory. In particular, we find the agreement in displacement statistics to be remarkably quantitative. We then examine numerical simulations of the TKdV deterministic dynamics, which elucidates the length and timescales required for statistical transitions in the experiments. Finally, this section discusses a peculiar power-law relationship between surface-displacement skewness and surface-slope variance predicted by the statistical mechanics framework \cite{majda2019}. New experimental measurements conclusively confirm this prediction, further demonstrating the predictive power of the TKdV framework developed by \maetal \cite{majda2019}. We close with some final remarks in Section \ref{conclusion}.

>>>>

\begin{abstract} 
Recent laboratory experiments of \boetal demonstrate that an abrupt change in bottom topography can trigger anomalous statistics in randomized surface waves. Motivated by these observations, \maetal developed a theoretical framework, based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) system, that successfully captures key qualitative features of the experiments, including the robust emergence of anomalous statistics and heightened skewness in the outgoing wavefield. Here, we extend these parallel experimental and modeling efforts with several new findings that have resulted from a synergetic interaction between the two. By precisely relating model parameters to physical ones, we calibrate the model inverse temperature to the specific conditions present in the experiments, thereby permitting a quantitative comparison. We find theoretically predicted distributions of surface displacement to match the experimental measurements with surprising detail. Prompted by the presence of surface slope in the TKdV Hamiltonian, we present new experimental measurements on surface slope statistics and compare them to model predictions. Analysis of some deterministic trajectories of TKdV elucidates the experimental length and time scales required for the statistical transition to a skewed state. Finally, the theory predicts a peculiar relationship between the outgoing displacement skewness and the change in slope variance, specifically how their ratio depends on the wave amplitude and depth ratio. New experimental measurements confirm this prediction in spectacular fashion.
%New experimental measurements provide convincing evidence for this prediction.
\end{abstract}

@@

Our particular focus is the laboratory experiments of \boetal~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing over the step, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, \maetal \cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda \& Qi (2019) \cite{majdaqi2019} analyzed more severely truncated systems --- as low as two interacting modes --- that exhibit a statistical phase transition to anomalous statistics and the creation of extreme events while enjoying a more tractable structure. More recently, Qi \& Majda (2019) \cite{MachineLearning2019} demonstrated the capability of machine learning strategies to accurately predict these extreme events. That work employs a deep neural network coupled with a judicious choice of the entropy loss function that emphasizes the dominant structures of the turbulent field \cite{MachineLearning2019}. It should be noted that, in these studies, extreme events are represented by strong skewness of the wave-field. They occur intermittently and on a relatively frequent basis, which is in contrast to the simpler situation of isolated, rare events \cite{guth2019machine}.

The purpose of the present manuscript is to provide a more comprehensive treatment of the {\em combined} experimental and theoretical efforts of \boetal~\cite{bolles2019} and \maetal~\cite{majda2019, majdaqi2019}. We present a variety of new findings that have resulted from a synergetic interaction between theory and experiments -- a cooperative strategy with a proven record of success \cite{camassa2012stratified, ristroph2012, ganedi2018equilibrium}. 
That outline of the paper is as follows. In Section \ref{experiments} we detail the laboratory experiments of \boetal \cite{bolles2019} and summarize previously reported findings on anomalous statistics of the surface displacement. We also present new experimental data on surface {\em slope} statistics --- a line of inquiry motivated by the theoretical advancements of \maetal \cite{majda2019}, specifically the central role played by the slope in the Hamiltonian structure of TKdV.  Section \ref{theory} discusses the TKdV theoretical framework, including both the deterministic and the statistical mechanics perspectives. In this section, we flesh out details of the non-dimensionalization, which, for brevity, were only briefly discussed in previous work. This exercise provides a more precise link between model and experimental parameters, ultimately facilitating a closer comparison between the two. We then discuss TKdV as a deterministic dynamical system, approachable from the viewpoint of statistical mechanics. The later perspective is based on the Hamiltonian structure and novel ensemble distributions that incorporate canonical and micro-canonical aspects \cite{abramov2003}.

Section \ref{results} presents a systematic comparison between experiments and theory, focusing on both surface displacement and surface slope. We find the statistical distributions of these two quantities to agree well across experiments and theory. In particular, we find remarkably quantitative agreement in the displacement statistics. We then examine numerical simulations of the TKdV deterministic dynamics, which elucidates the length and timescales required for statistical transitions in the experiments. Finally, this section discusses a peculiar power-law relationship between surface-displacement skewness and surface-slope variance predicted by the statistical mechanics framework \cite{majda2019}. New experimental measurements conclusively confirm this prediction, further demonstrating the predictive power of the TKdV framework developed by \maetal \cite{majda2019}. We close with some final remarks in Section \ref{conclusion}.
